\vspace{-2.0em} Move Schedules: Fast persistence computations in coarse dynamic settings \thanks{ \textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758. }

### 2023-01-28 move_schedules.tex
Greatly simplified natural images section, finishing the simplifications of the multiparameter use case

\subsection*{Multiparameter persistence}
Given a procedure to filter a space in multiple dimensions simultaneously, a \emph{multifiltration}, the goal of multi-parameter persistence is to identify persistent features by examining the entire multifiltration. 
Such a generalization has appeared naturally in many application contexts, showing potential as a tool for exploratory data analysis~\cite{lesnick2012multidimensional}. Indeed, one of the drawbacks of persistence is its instability with respect to strong outliers, which can obscure the detection of significant topological structures~\cite{buchet2015topological}.
One exemplary use case of multi-parameter persistence is to detect these strong outliers by filtering the data with respect to both density and the associated metric.

Unfortunately, unlike the one-parameter case, there is no complete discrete invariant for multi-parameter persistence.
Circumventing this, Lesnick et al~\cite{lesnick2015interactive} associate a variety of incomplete invariants to 2-parameter persistence modules; we focus here on the \emph{fibered barcode} invariant, defined as follows: 
\begin{definition}[Fibered barcode]
	%\in \overline{\mathcal{L}}
	The fibered barcode $\mathcal{B}(M)$ of a 2D persistence module $M$ is the map which sends each line  $L \subset \mathbb{R}^2$ with non-negative slope to the barcode $\mathcal{B}_L(M)$: 
%	$$ L \mapsto \mathcal{B}_L(M) $$
$$ \mathcal{B}(M) = \{ \; B_L(M) : L \in \mathbb{R} \times \mathbb{R}^{+} \; \}$$
Equivalently, $\mathcal{B}(M)$ is the 2-parameter family of barcodes given by restricting $M$ to the of set affine lines with non-negative slope in $\mathbb{R}^2$. 
\end{definition} 
\noindent Although an intuitive invariant, it is not clear how one might go about computing $\mathcal{B}(M)$ efficiently. 
One obvious choice is fix $L$ via a linear combination of two filter functions, restrict $M$ to $L$, and compute the associated 1-parameter barcode. 
However, this is an $O(m^3)$ time computation, which is prohibitive for exploratory data analysis purposes.

@@

 \subsubsection*{Natural images dataset} 
As we have outlined the computational theory of 2-parameter persistence and elucidated it relevance to our proposed move scheduling approach, we now demonstrate the efficiency of scheduling with a practical application: detecting the presence of a low-dimensional topological space which well-approximates the distribution of patches from high-contrast images.

A common hypothesis is that high dimensional data tend to lie in the vicinity of an embedded, low dimensional manifold or topological space. An exemplary demonstration of this is given in~\cite{lee2003nonlinear}, who explored the space of $3 \times 3$ high-contrast patches extracted from Hans van Hateren's still image collection~\cite{hateren_schaaf_1998}, which consists of approximately 4,000 monochrome images\footnote{See \url{http://bethgelab.org/datasets/vanhateren/} for details on the image collection.} depicting various areas outside around Groningen (Holland). Motivated by applications to sparse coding applications, Lee et al.~\cite{lee2003nonlinear} was interested in how these natural image patches were distributed, in pixel-space, with respect to predicted spaces and manifolds. One of the motivating questions of their research was whether there were any clear qualitative differences the distributions of patches extracted from images of different modalities, such optical and range images. A major finding from their work is that the majority of high-contrast image patches are concentrated around a 2-dimensional submanifold embedded in $S^7$.
Supplemental experiments conducted by Carllsson et al.~\cite{carlsson2008local} using persistent homology confirmed that the distribution of high-contrast $3 \times 3$ patches is indeed well-approximated by a Klein bottle---around 60\% of the high-contrast patches from the still image data set are concentrated within a small neighborhood around a model of a Klein bottle which accounts for $\approx$ 21\% of the volume $S^7$, compared to the $\approx$ 84\% of volume of $S^7$ accounted for by the entire data set. Along a similar vein, in the context of sparse coding, Perea et al.~\cite{perea2014klein} introduced a dictionary learning framework for estimating and representing the distribution of patches from texture images.

@@

If one was not aware of the analysis done by~\cite{lee2003nonlinear, hateren_schaaf_1998, carlsson2008local, perea2014klein}, it is not immediately clear a priori that the Klein bottle model is indeed a good candidate for capturing non-linearity of image patches. Nonetheless, as we have mentioned, among the first uses of persistent homology was the so-called ``homology inference problem''~\cite{perea2018brief}; given a suitable treatment of strong outliers using by e.g. filtering both on the geometry and the density of the image patches, the homology of the Klein bottle should become apparent through a persistence computation.

Consider the (coarsened) fibered barcode computed from a standard Rips / codensity bifiltration on a representative sample of the image data from~\cite{hateren_schaaf_1998}, shown in Figure~\ref{fig:patch_data_dgm}. 
From the bigraded Betti number and the dimension function, one finds that a large area of dimension function is constant (highlighted as the blue portion in the middle of Figure~\ref{fig:patch_data_dgm}), wherein the first Betti number is 5. There are many spaces whose  meets this criterion, though upon further inspection, one candidate space that seems particularly plausible is the three-circle model $C_3$. The $C_3$ model consists of three circles, two of which (say, $S_v$ and $S_h$) intersect the third (say, $S_\mathrm{lin}$) in exactly two points, but themselves do not intersect. Upon performing an analysis on the density of the patches analogous to the one done by~\cite{lee2003nonlinear}, one finds that the mean-centered and contrast normalized $3 \times 3$ patches like on a 7-dimensional ellipsoid $\tilde{S}^7 \subset \mathbb{R}^9$. Contrast is typically measured using a discrete version of the Dirichlet semi-norm, or the ``$D$-norm'' $\lVert \cdot \rVert_D$ for short, which is a unique scale-invariant norm on images:
$$ \lVert x \rVert_D = \sqrt{\sum_{i \sim j}(x_i - x_j)^2} = \sqrt{x^T D x}$$
where $D$ is a fixed matrix which upon application to an image $x \in \mathbb{R}^9$ yields a value proportional to the sum of the differences between the 4-connected neighbors (described by the relation $i \sim j$). It is advantageous to ``whiten'' the data via a change of coordinates to Euclidean sphere. 
This can be achieved by expressing the patch data as an expansion of a certain convenient basis of 8 non-constant eigenvectors, often called the Discrete Cosine Transform (DCT) basis~\cite{lee2003nonlinear}, shown below: 
%\vspace{-0.2em}
\begin{center}
	\includegraphics[width=0.80\textwidth]{dct_basis} 
\end{center}
%\begin{figure}[h]
%	\captionsetup{labelformat=empty}
%	\centering
%	\includegraphics[width=0.90\textwidth]{dct_basis}
%	\caption{}
%\end{figure}
%\vspace{-1.5em}
\noindent Projecting the image data onto the first two basis vectors of leads to the projection shown in the top left of Figure~\ref{fig:patch_data_dgm}. 15 landmark points are also shown. Observe the data are distributed well around three ``circles''---the outside circle capturing the rotation gradient of the image patches ($S_{\mathrm{lin}}$), and the other two circles capturing the vertical and horizontal gradients ($S_v$ and $S_h$, respectively). Since the three circle model is the 1-skeleton of the Klein bottle~\cite{carlsson2008local}, one may conclude that the Klein bottle might be a reasonable candidate upon which the image data are distributed. This can be further verified empirically by e.g. using the iterative, meanshift-like procedure described in~\cite{carlsson2008local} to fit a parameterization of the model in question to the data.

Using the same high-contrast patch data set studied in~\cite{lee2003nonlinear}, we evaluate the performance of various methods at computing the fibered barcode invariant using the parameterization from~\ref{sec:fibered_barcode_reparam}. 
Due to the aformentioned high complexity of the fibered barcode computation, we begin by working with a subset of the image patch data $\mathcal{X}$. Specifically, we combine the use furthest-point sampling and proportionate allocation (stratified) sampling to sample landmarks $X \subset \mathcal{X}$ distributed within $n = 25$ strata. Each strata consists of the $(1/n)$-thick level set given by $\rho_{15}$, the density estimator. The use of furthest-point sampling gives us certain coverage guarantees that the geometry is approximately preserved within each level set, whereas the stratification ensures the original density of is approximated preserved as well. From this data set, we construct a Rips-(co)density bifiltration using $\rho_{15}$ equipped with the geodesic metric computed over the same $k = 15$ knn graph on $X$. 

@@

\noindent
As shown on the table, when the coarsening $\kappa$ is small enough, we're able to achieve a significant reduction in the number of total column operations needed to compute $\mathcal{T}$ compared to both vineyards and $\mathtt{pHcol}$. %though it's worth noting that permutations are constant time for vineyards while they require at most $O(m)$ elementary operations for moves. 
This is further reinforced by the observation that, when there is a high degree of coarsening, vineyards is particularly inefficient and moves requires only about 3x less column operations that naively computing $\mathcal{T}$ independently. As the coarsening becomes more refined and more 2-cells are added to $\mathcal{A}(M)$, however, vineyards quickly becomes a much more viable option compared to $\mathtt{pHcol}$---as predicted---though even at the highest coarsening we tested the gain in efficiency is relatively small. In contrast, our proposed moves approach scales quite well with this refinement, requiring about $12\%$  and $\approx 5\%$ of the number of column operations as vineyards and $\mathrm{pHcol}$, respectively.

>>>>

\subsection{Multiparameter persistence}
Given a procedure to filter a space in multiple dimensions simultaneously, a \emph{multifiltration}, the goal of multi-parameter persistence is to identify persistent features by examining the entire multifiltration. 
Such a generalization has appeared naturally in many application contexts, showing potential as a tool for exploratory data analysis~\cite{lesnick2012multidimensional}. Indeed, one of the drawbacks of persistence is its instability with respect to strong outliers, which can obscure the detection of significant topological structures~\cite{buchet2015topological}.
One exemplary use case of multi-parameter persistence is to detect these strong outliers by filtering the data with respect to both density and the associated metric.
In this section, we show the utility of scheduling with a real-world use case: detecting the presence of a low-dimensional topological space which well-approximates the distribution of natural images. 
As a quick outline, in what follows we briefly recall the fibered barcode invariant~\ref{sec:fibered_barcode}, summarize its potential application to a particular data set with known topological structure~\ref{sec:natural_images}, and conclude with experiments of demonstrating how scheduling enables such applications~\ref{sec:empirical_klein}.

\subsubsection{Fibered barcode}\label{sec:fibered_barcode}
Unfortunately, unlike the one-parameter case, there is no complete discrete invariant for multi-parameter persistence.
Circumventing this, Lesnick et al~\cite{lesnick2015interactive} associate a variety of incomplete invariants to 2-parameter persistence modules; we focus here on the \emph{fibered barcode} invariant, defined as follows: 
\begin{definition}[Fibered barcode]
	%\in \overline{\mathcal{L}}
	The fibered barcode $\mathcal{B}(M)$ of a 2D persistence module $M$ is the map which sends each line  $L \subset \mathbb{R}^2$ with non-negative slope to the barcode $\mathcal{B}_L(M)$: 
%	$$ L \mapsto \mathcal{B}_L(M) $$
$$ \mathcal{B}(M) = \{ \; B_L(M) : L \in \mathbb{R} \times \mathbb{R}^{+} \; \}$$
Equivalently, $\mathcal{B}(M)$ is the 2-parameter family of barcodes given by restricting $M$ to the of set affine lines with non-negative slope in $\mathbb{R}^2$. 
\end{definition} 
\noindent Although an intuitive invariant, it is not clear how one might go about computing $\mathcal{B}(M)$ efficiently. 
One obvious choice is fix $L$ via a linear combination of two filter functions, restrict $M$ to $L$, and compute the associated 1-parameter barcode. 
However, this is an $O(m^3)$ time computation, which is prohibitive for exploratory data analysis purposes.

@@

 \subsubsection{Natural images dataset}\label{sec:natural_images}
A common hypothesis is that high dimensional data tend to lie in the vicinity of an embedded, low dimensional manifold or topological space. An exemplary demonstration of this is given in the analysis by Lee et al.~\cite{lee2003nonlinear}, who explored the space of high-contrast patches extracted from Hans van Hateren's~\cite{hateren_schaaf_1998} still image collection\footnote{See \url{http://bethgelab.org/datasets/vanhateren/} for details on the image collection.}, which consists of $\approx 4,000$ monochrome images depicting various areas outside Groningen (Holland). 
In particular,~\cite{lee2003nonlinear} were interested in exploring how high-contrast $3 \times 3$ image patches  were distributed, in pixel-space, with respect to predicted spaces and manifolds.
Formally, they measured contrast using a discrete version of the scale-invariant Dirichlet semi-norm:
$$ \lVert x \rVert_D = \sqrt{\sum_{i \sim j}(x_i - x_j)^2} = \sqrt{x^T D x}$$
where $D$ is a fixed matrix which upon application to an image $x \in \mathbb{R}^9$ yields a value proportional to the sum of the differences between each pixels 4 connected neighbors (described above by the relation $i \sim j$).
Their research was primarily motivated by discerning whether there existed clear qualitative differences in the distributions of patches extracted from images of different modalities, such optical and range images.
By mean-centering, contrast normalizing, and``whitening'' the data via the Discrete Cosine Transform (DCT) basis~\cite{lee2003nonlinear}, they found they could form a convenient basis for $D$ via an expansion of 8 certain non-constant eigenvectors, shown below: 
%\vspace{-0.2em}
\begin{center}
	\includegraphics[width=0.80\textwidth]{dct_basis} 
\end{center}
%\begin{figure}[h]
%	\captionsetup{labelformat=empty}
%	\centering
%	\includegraphics[width=0.90\textwidth]{dct_basis}
%	\caption{}
%\end{figure}
%\vspace{-1.5em}
Since these images are scale-invariant, the expansion of these basis vectors spans the 7-sphere, $S^7 \subset \mathbb{R}^8$. Using a voronoi cell decomposition of the data, their subsequent distribution analysis suggested that the majority of data points concentrated in a few high-density regions.

After Lee et al published their work, Carlsson et al.~\cite{carlsson2008local} subsequently performed extensive experiments conducted using persistent homology, wherein he found that the distribution of high-contrast $3 \times 3$ patches is actually well-approximated by a Klein bottle $\mathcal{M}$---around 60\% of the high-contrast patches from the still image data set lie within a small neighborhood around $\mathcal{M}$ which accounts for only 21\% of the volume $S^7$. Along a similar vein, in the context of sparse coding, Perea et al.~\cite{perea2014klein} introduced a dictionary learning framework for estimating and representing the distribution of patches from texture images.

@@

If one was not aware of the analysis done by~\cite{lee2003nonlinear, hateren_schaaf_1998, carlsson2008local, perea2014klein}, it is not immediately clear a priori that the Klein bottle model is  a good candidate for capturing non-linearity of image patches. 
Indeed, armed with fine topological intuition, Carlsson still needed to perform extensive sampling, preprocessing, and model fitting techniques in order reveal the underlying the topological space with persistent homology~\cite{carlsson2008local}.
One reason such preprocessing is needed is due to persistent homology's aforementioned instability with respect to strong outliers. 
In the ideal setting, a multi-parameter approach that accounts for the local density of points should require far less experimentation. 
%Nonetheless, given a suitable treatment of strong outliers using by e.g. filtering both on the geometry and the density of the image patches, the homology of the Klein bottle should become apparent through a persistence computation.
%Nonetheless, as we have mentioned, among the first uses of persistent homology was the so-called ``homology inference problem''~\cite{perea2018brief}; given a suitable treatment of strong outliers using by e.g. filtering both on the geometry and the density of the image patches, the homology of the Klein bottle should become apparent through a persistence computation.

Consider the (coarsened) fibered barcode computed from a standard Rips / codensity bifiltration on a representative sample of the image data from~\cite{hateren_schaaf_1998}, shown in Figure~\ref{fig:patch_data_dgm}. 
From the bigraded Betti number and the dimension function, one finds that a large area of dimension function is constant (highlighted as the blue portion in the middle of Figure~\ref{fig:patch_data_dgm}), wherein the first Betti number is 5. Further inspection suggests one plausible candidate is the three-circle model $C_3$, which consists of three circles, two of which (say, $S_v$ and $S_h$) intersect the third (say, $S_\mathrm{lin}$) in exactly two points, but themselves do not intersect. 
Projecting the image data onto the first two basis vectors of leads to the projection shown in the top left of Figure~\ref{fig:patch_data_dgm}, of which 15 landmark points are also shown. Observe the data are distributed well around three ``circles''---the outside circle capturing the rotation gradient of the image patches ($S_{\mathrm{lin}}$), and the other two capturing the vertical and horizontal gradients ($S_v$ and $S_h$, respectively). Since the three circle model is the 1-skeleton of the Klein bottle, one may concur with Carlssons analysis~\cite{carlsson2008local} that the Klein bottle may be a reasonable candidate upon which the image data are distributed. 
% This can be further verified empirically by e.g. using the iterative, meanshift-like procedure described in~\cite{carlsson2008local} to fit a parameterization of the model in question to the data. 
The degree to which multi-parameter persistence simplifies this exploratory phase cannot be understated: we believe multi-parameter persistence has a larger role to play in manifold learning. 
Unfortunately, as mentioned prior, the compute barriers effective bar its use in practice.

\subsubsection{Accelerating 2D persistence}\label{sec:empirical_klein}
As we have outlined the computational theory of 2-parameter persistence and elucidated its relevance to our proposed move scheduling approach, we now demonstrate the efficiency of scheduling using the same high-contrast patch data set studied in~\cite{lee2003nonlinear} by evaluating the performance of various methods at computing the fibered barcode invariant via the parameterization from~\ref{sec:fibered_barcode_reparam}.

Due to the aforementioned high complexity of the fibered barcode computation, we begin by working with a subset of the image patch data $\mathcal{X}$. 
In particular, we combine the use of furthest-point sampling and proportionate allocation (stratified) sampling to sample landmarks $X \subset \mathcal{X}$ distributed within $n = 25$ strata. Each strata consists of the $(1/n)$-thick level set given by $\rho_{15}$, the density estimator. The use of furthest-point sampling gives us certain coverage guarantees that the geometry is approximately preserved within each level set, whereas the stratification ensures the original density of is approximated preserved as well. 
From this data set, we construct a Rips-(co)density bifiltration using $\rho_{15}$ equipped with the geodesic metric computed over the same $k = 15$ k-nearest neighbor graph on $X$. 

@@

As shown on the table, when the coarsening $\kappa$ is small enough, we're able to achieve a significant reduction in the number of total column operations needed to compute $\mathcal{T}$ compared to both vineyards and $\mathtt{pHcol}$. %though it's worth noting that permutations are constant time for vineyards while they require at most $O(m)$ elementary operations for moves. 
This is further reinforced by the observation that, when there is a high degree of coarsening, vineyards is particularly inefficient and moves requires only about 3x less column operations that naively computing $\mathcal{T}$ independently. As the coarsening becomes more refined and more 2-cells are added to $\mathcal{A}(M)$, however, vineyards quickly becomes a much more viable option compared to $\mathtt{pHcol}$---as predicted---though even at the highest coarsening we tested the gain in efficiency is relatively small. In contrast, our proposed moves approach scales quite well with this refinement, requiring about $12\%$  and $\approx 5\%$ of the number of column operations as vineyards and $\mathrm{pHcol}$, respectively.
